---
title: "Sequential Block Merging Algorithm for General Isotone Optimization"
author: |
  | Xiwen Wang and Daniel P. Palomar
  | The Hong Kong University of Science and Technology (HKUST)
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: vignette
    toc: yes
    toc_depth: 2
  pagedown::html_paged:
csl: ieee.csl
bibliography: refs.bib 
vignette: >
  %\VignetteIndexEntry{SBM_Isotone}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, echo=FALSE}
library(knitr)
opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.retina = 2,
  out.width = "85%",
  dpi = 96 
) 
options(width=100)
# rmarkdown::render("vignettes/DesignOfHighOrderPortfolios.Rmd", "prettydoc::html_pretty")
```

# Quick Start

Let's first load the package:

```{r setup}
library(IsotoneOptimization)
```

# Solving Isotonic Regression under chain/binary tree/2d-grid isotonic constraints

The package __IsotoneOptimization__ implements the Sequential Block Merging algorithm [@wang2022efficient] for solving isotonic regression under constraints that can be represented by arbitrary directed acyclic graph. Here are some structures which are frequently used. 

## Chain isotonic constraints

Given optimization variable $\mathbf{x}\in \mathbb{R}^n$, the isotonic regression under chain isotonic constraints are given as 

$$
\begin{array}{rl} 
\underset{\mathbf{x}}{\mathsf{minimize}} & \frac{1}{2}\left\|\mathbf{x}-\mathbf{y}\right\|_2^2 \\
\mathsf{subject}\text{ }\mathsf{to} & x_1\leq x_2\leq \cdots\leq x_n.
\end{array}
$$
```{r, cache=TRUE}
y <- 1:100 /10 + rnorm(100, 0, 3)
y_calibrated <- solve_isotone_chain(y)

plot(1:100,y,type="o",col="red")
lines(1:100,y_calibrated,col="blue")
```

## Binary tree isotonic constraints

Given optimization variable $\mathbf{x}\in \mathbb{R}^n$, the isotonic regression under binary isotonic constraints are given as 

$$
\begin{array}{rl} 
\underset{\mathbf{x}}{\mathsf{minimize}} & \frac{1}{2}\left\|\mathbf{x}-\mathbf{y}\right\|_2^2 \\
\mathsf{subject}\text{ }\mathsf{to} & x_i \leq x_{2i} ,i=1,\dots,\left\lceil \frac{n}{2}\right\rceil,\\
 & x_i\leq x_{2i+1},i=1,\dots,\left\lceil \frac{n-1}{2}\right\rceil.
\end{array}
$$
We can solve the problem and plot the results via package igraph.

```{r, cache=TRUE}
library(igraph)
p <- 20
# build a tree graph
tr <- make_tree(p, children = 2, mode = "out")
y <- 1:20 /10 + rnorm(20, 0, 3)
y_calibrated <- solve_isotone_binary_tree(y)

V(tr)$name <- as.character(round(y_calibrated,2))
plot(tr,  vertex.size = 4,          # Smaller nodes
    vertex.label = V(tr)$name, # Set the labels
    vertex.label.cex = 1,   # Slightly smaller font
    vertex.label.dist = 1.4,  # Offset the labels
    vertex.label.color = "black")
```

## 2d-grid isotonic constraints

Given optimization variable $\mathbf{x}\in \mathbb{R}^n$, the isotonic regression under binary isotonic constraints are given as 

$$
\begin{array}{rl} 
\underset{\mathbf{x}}{\mathsf{minimize}} & \frac{1}{2}\left\|\mathbf{x}-\mathbf{y}\right\|_2^2 \\
\mathsf{subject}\text{ }\mathsf{to} & x_{i+(j-1)\times\left\lfloor\sqrt{n}\right\rfloor} \leq x_{i+1+(j-1)\times\left\lfloor\sqrt{n}\right\rfloor} ,\forall i=1,\dots,\left\lfloor\sqrt{n}\right\rfloor -1,j=1,\dots,\left\lfloor\sqrt{n}\right\rfloor\\
 & x_{i+(j-1)\times\left\lfloor\sqrt{n}\right\rfloor} \leq x_{i+j\times\left\lfloor\sqrt{n}\right\rfloor} ,\forall i=1,\dots,\left\lfloor\sqrt{n}\right\rfloor ,j=1,\dots,\left\lfloor\sqrt{n}\right\rfloor-1,\\
 & x_i\leq x_{i+1},\forall i=\left\lfloor\sqrt{n}\right\rfloor^2,\dots, n-1.
\end{array}
$$

We can solve the problem and plot the results via package igraph.

```{r, cache=TRUE}
library(igraph)
p <- 25
sqrt_p <- 5
E_mat <- c()
# build a 2d-grid graph
for(i in 1:4)
  for(j in 1:5) 
    E_mat <- cbind(E_mat, c(i+(j-1)*sqrt_p, i+1+(j-1)*sqrt_p))

for(i in 1:5)
  for(j in 1:4) 
    E_mat <- cbind(E_mat, c(i+(j-1)*sqrt_p, i+j*sqrt_p))

E_mat <- t(E_mat)

y <- 1:p /10 + rnorm(p, 0, 3)
y_calibrated <- solve_isotone_2d_grid(y)

grid_graph <- graph_from_edgelist(E_mat)
V(grid_graph)$name <- as.character(round(y_calibrated,2))
plot(grid_graph,  vertex.size = 4,          # Smaller nodes
    vertex.label = V(grid_graph)$name, # Set the labels
    vertex.label.cex = 1,   # Slightly smaller font
    vertex.label.dist = 1.4,  # Offset the labels
    vertex.label.color = "black")
```

# Solving Isotonic Regression with isotonic constraints via arbitrary DAG

We can also define any constraints graph as long as it is directed acyclic. The below example generate a random DAG. By using the function __Ordered_DAG__, we solve the following problem 

$$
\begin{array}{rl} 
\underset{\mathbf{x}}{\mathsf{minimize}} & \frac{1}{2}\left\|\mathbf{x}-\mathbf{y}\right\|_2^2 \\
\mathsf{subject}\text{ }\mathsf{to} & x_i \leq x_{j},\forall (i,j)\in \mathcal E.
\end{array}
$$

```{r, cache=TRUE}
set.seed(12)
start_nodes <- rep(1:10,3)
end_nodes <- start_nodes + round(runif(30, 1, 5))

E_mat <- cbind(start_nodes, end_nodes)

DAG_graph <- graph_from_edgelist(E_mat)  
DAG_graph <- simplify(
    DAG_graph,
    remove.multiple = TRUE,
    remove.loops = TRUE,
    edge.attr.comb = igraph_opt("edge.attr.comb")
)

E_mat <- as_edgelist(DAG_graph)

y <- 1:max(E_mat)/2 + rnorm(max(E_mat), 0, 3)
y_calibrated <- solve_isotone_DAG(y, rep(1,max(E_mat)), E_mat)

V(DAG_graph)$name <- as.character(round(y_calibrated,2))
plot(DAG_graph,  vertex.size = 4,          # Smaller nodes
    vertex.label = V(DAG_graph)$name, # Set the labels
    vertex.label.cex = 1,   # Slightly smaller font
    vertex.label.dist = 1.4,  # Offset the labels
    vertex.label.color = "black")
```



# References {-}

\setlength{\parindent}{-0.2in}
\setlength{\leftskip}{0.2in}
\setlength{\parskip}{8pt}
\noindent


